We presented a model-agnostic posterior sampling method for data assimilation with flow-based generative models. The unifying object is the affine Gaussian probability path: conditioning it on an observation tilts its target to the posterior while preserving the path structure, so the conditional score and the conditional velocity correct the prior ones by a single likelihood-score term, scaled respectively by one and by the velocity--score duality coefficient $\vscoef_\tau$. Applying the marginal-preserving family of SDEs to this conditioned path yields a one-parameter family of posterior samplers whose three members of practical interest -- the stochastic-interpolant SDE, the flow matching SDE, and a deterministic guided probability-flow ODE -- share a single guidance weight $\gweight_\tau=\vscoef_\tau+\tfrac12\gdiff_\tau^2$ and differ only in the diffusion coefficient. A tractable interpolated-observation likelihood supplies the likelihood score in closed form, and a closed-form multiplicative gain relates it to the Gaussian-surrogate likelihood score at the cost of a single rank-$N_y$ correction per step, identically for all three samplers.

\paragraph{Limitations.}
Several assumptions delimit the exactness guarantees. First, the closed-form likelihood requires a linear observation operator and Gaussian observation noise; nonlinear operators would require linearisation or sampling-based estimates. Second, the interpolated-observation likelihood and the true-likelihood surrogate in Section~\ref{sec:multiplicative_correction} are Gaussian approximations whose conditional moments are exact but whose shape is not, except in the linear-Gaussian setting; the multiplicative correction is exact only relative to these surrogates. Third, autoregressive one-step assimilation approximates the trajectory posterior: it omits the importance weights $\conddist{\obs^{n}}{\state^{n-1}}$ and the conditioning on future observations (Section~\ref{sec:preliminaries}). Fourth, the deterministic guided ODE is unavailable for stochastic interpolants, whose point-mass source carries no randomness at zero diffusion; it is well posed only for flow matching, where the source $\mathcal{N}(0,\bI)$ is non-degenerate. (For the marginal construction the flow matching initialisation needs no reweighting: $\Phi_0^{\obs}$ is constant at $\tau=0$, so $\mathcal{N}(0,\bI)$ is exact; reweighting by $\Phi_0^{\obs}$ is required only for the path-space Doob sampler of Appendix~\ref{appendix:proof_posterior}.) Finally, all guarantees are stated for a perfectly trained drift or velocity; quantifying the effect of model error on the posterior remains an open question, which we aim to address in future work.
