We denote discrete physical time steps with superscripts and pseudo-time steps with subscripts. E.g. $\state^{n}$ denotes a quantity at physical time step $n$ and $\bx_{\tau}$ denotes a quantity at pseudo-time $\tau$. Furthermore, the notation $\state^{n:m}$, with $n < m$, refers to the time series $(\state^{n}, \state^{n+1}, \ldots, \state^{m-1}, \state^{m})$. We write $X\sim \dist{x}$ when $X$ is distributed according to $P(X)$ and $X\sim \conddist{x}{y}$ when $X$ is distributed according to $P(X|Y)$, identifying distributions through their densities.

\subsection{Bayesian data assimilation}
We briefly outline the Bayesian data assimilation setting in this subsection. For more information on data assimilation, see \cite{carrassi_data_2018, evensen_data_2022}.

We consider the discrete dynamical system:
\begin{subequations} \label{eq:data_assimilation_problem}
\begin{align}
    \state^{n} &= \forwardmodel(\state^{n-1}) + \modelnoise^{n},  &\modelnoise^{n}\sim \dist[\modelnoise], \label{eq:forward}\\
    \obs^{n} &= \obsoperator(\state^{n}) + \obsnoise^{n},  &\obsnoise^{n}\sim \dist[\obsnoise], \label{eq:observation}
\end{align}
\end{subequations}
where, for all $n=1,\ldots,N$, $\state^{n}\in\R^{\statedim}$ is the state, $\obs^n\in\R^{\obsdim}$ are the observations, $\modelnoise^n\in\R^{\statedim}$ is the model noise, $\obsnoise^n\in\R^{\obsdim}$ is the observation noise, $\forwardmodel:\R^{\statedim}\rightarrow\R^{\statedim}$ is the forward operator, and $\obsoperator:\R^{\statedim}\rightarrow\R^{\obsdim}$ is the observation operator. The system can be the result of a discretized PDE system and is therefore potentially very high-dimensional. Eq. \eqref{eq:forward} induces a prior transition density, $\conddist{\state^{n}}{\state^{n-1}}$, that propagates the state without incorporating observations. Furthermore, the observation likelihood can be evaluated via the noise distribution, $\conddist{\obs^n}{\state^{n}} = \dist[\eta]{\obs^n-\obsoperator(\state^{n})}$.

In this work, we are interested in the posterior distribution over trajectories, which by the Markovian property of Eq.~\eqref{eq:data_assimilation_problem} factorizes as \cite{carrassi_data_2018, evensen_data_2022}
\begin{align} \label{eq:posterior_trajectory_distribution}
    \conddist{\state^{0:n}}{\obs^{1:n}} \propto \dist{\state^{0}} \prod_{i=1}^{n} \conddist{\obs^{i}}{\state^{i}}\,\conddist{\state^{i}}{\state^{i-1}}.
\end{align}
We approximate samples from this distribution by autoregressively sampling from the one-step posterior:
\begin{align} \label{eq:posterior_distribution}
    \conddist{\state^{n}}{\obs^n, \state^{n-1}} = \frac{\conddist{\obs^{n}}{\state^{n}}\conddist{\state^{n}}{\state^{n-1}}}{\conddist{\obs^{n}}{\state^{n-1}}}.
\end{align}
This sequential scheme is exact for a single assimilation step. Over multiple steps it corresponds to propagating with the optimal proposal of particle filtering while omitting the associated importance weights $\conddist{\obs^{i}}{\state^{i-1}}$ and the conditioning on future observations, an approximation that is standard in autoregressive generative data assimilation \cite{rozet_score-based_2023, chen_flowdas_2025}. We therefore focus on sampling from Eq. \eqref{eq:posterior_distribution} for the rest of the paper. The prior $\conddist{\state^{n}}{\state^{n-1}}$ is provided by a flow-based generative model trained on forecast data, which we introduce next in a model-agnostic way.

\subsection{Flow-based generative models as affine Gaussian probability paths}
\label{subsec:flow_models}

Stochastic interpolants and flow matching are usually presented separately, the former as an SDE sampler and the latter as an ODE sampler. We treat them uniformly as instances of an \emph{affine Gaussian probability path}, a unification established in prior work \cite{albergo_stochastic_2023, lipman_flow_2023, ma_sit_2024, holderrieth_generator_2024} and all that the proposed data assimilation method requires. The goal is to build a generative model that samples from a target conditional density $\conddist{\bx_1}{\bx_0}$ given a sample $\bx_0$; in the forecasting setting we set $\bx_0 = \state^{n-1}$ and $\bx_1 = \state^{n}$ to recover the prior in Eq.~\eqref{eq:posterior_distribution}. Sampling is carried out over a pseudo-time $\tau\in[0,1]$ that is unrelated to physical time and serves only to transport a tractable source distribution to the target.

\paragraph{Affine Gaussian interpolant.}
The common building block is a stochastic process that interpolates, in pseudo-time, between a tractable source and the target,
\begin{align}\label{eq:general_interpolant}
    \bx_\tau = \underbrace{\alpha_\tau \bm{a}_0 + \sigpath_\tau \latentnoise}_{\text{source contribution}} + \beta_\tau \bx_1,
    \qquad \latentnoise \sim \mathcal{N}(0,\bI),
\end{align}
where $\bm{a}_0$ is a deterministic anchor available at sampling time, $\latentnoise$ is a standard normal driver independent of $\bx_1$, and the scalar schedules $\alpha_\tau,\beta_\tau,\sigpath_\tau$ satisfy the boundary conditions $\beta_0=0$, $\beta_1=1$, and $\alpha_1=\sigpath_1=0$. These conditions guarantee that $\bx_\tau$ at $\tau=1$ equals the target $\bx_1$ and at $\tau=0$ equals the source $\alpha_0\bm a_0 + \sigpath_0\latentnoise$. Conditioned on $\bx_1$ (and on $\bm{a}_0$), Eq.~\eqref{eq:general_interpolant} is Gaussian with mean $\alpha_\tau\bm{a}_0 + \beta_\tau\bx_1$ and isotropic covariance $\sigpath_\tau^2\bI$; the marginal $p_\tau(\bx\mid\bx_0)$ is the corresponding Gaussian mixture over $\bx_1$. Two quantities are associated with this path.

The \emph{velocity field} is the conditional expectation of the instantaneous interpolant displacement,
\begin{align}\label{eq:general_velocity}
    \fmvel_\tau(\bx,\bx_0)
    = \E\!\left[\dot\alpha_\tau\bm{a}_0 + \dot\sigpath_\tau\latentnoise + \dot\beta_\tau\bx_1 \;\middle|\; \bX_\tau=\bx\right],
\end{align}
and is the drift of the probability-flow ODE $\rd\bX_\tau = \fmvel_\tau(\bX_\tau,\bx_0)\,\rd\tau$ that transports $p_0$ to $p_1$ \cite{lipman_flow_2023, albergo_stochastic_2023}. The \emph{score} is $\genscore_\tau(\bx,\bx_0) = \nabla_{\bx}\log p_\tau(\bx\mid\bx_0)$. For affine Gaussian paths the velocity and the score obey an exact linear identity. Introducing the \emph{velocity--score coefficient}
\begin{align}\label{eq:vscoef_def}
    \vscoef_\tau := \sigpath_\tau\!\left(\frac{\dot\beta_\tau}{\beta_\tau}\sigpath_\tau - \dot\sigpath_\tau\right)
    = \frac{\sigpath_\tau(\dot\beta_\tau\sigpath_\tau - \dot\sigpath_\tau\beta_\tau)}{\beta_\tau},
\end{align}
the velocity is an affine function of the state with the score entering linearly,
\begin{align}\label{eq:general_velocity_of_score}
    \fmvel_\tau(\bx)
    = \frac{\dot\beta_\tau}{\beta_\tau}\bx
    + \left(\dot\alpha_\tau - \frac{\dot\beta_\tau\alpha_\tau}{\beta_\tau}\right)\bm a_0
    + \vscoef_\tau\,\genscore_\tau(\bx),
\end{align}
which inverts to
\begin{align}\label{eq:general_score_velocity}
    \genscore_\tau(\bx)
    = \frac{\beta_\tau\,\fmvel_\tau(\bx) - \dot\beta_\tau\,\bx - (\dot\alpha_\tau\beta_\tau - \dot\beta_\tau\alpha_\tau)\,\bm a_0}
           {\beta_\tau\,\vscoef_\tau},
\end{align}
derived in Appendix~\ref{appendix:score_velocity}, so either object determines the other. In particular, the score is available in closed form from a model trained to produce the velocity, and vice versa. The coefficient $\vscoef_\tau$ is the only ingredient, beyond the velocity and the score themselves, that the deterministic posterior sampler of Section~\ref{sec:methodology} requires.

\paragraph{Lifting to a family of SDEs.}
A central fact we exploit is that the probability-flow ODE is only one member of a one-parameter family of dynamics with identical marginals \cite{song_score-based_2021, ma_sit_2024}. For any diffusion schedule $\gdiff_\tau\ge 0$, the SDE
\begin{align}\label{eq:general_sde_family}
    \rd\bX_\tau
    = \left[\fmvel_\tau(\bX_\tau,\bx_0) + \tfrac{1}{2}\gdiff_\tau^2\,\genscore_\tau(\bX_\tau,\bx_0)\right]\rd\tau
    + \gdiff_\tau\,\rd\bW_\tau, \qquad \bX_0\sim p_0,
\end{align}
has the same marginals $p_\tau(\cdot\mid\bx_0)$ as the ODE for every choice of $\gdiff_\tau$ (Appendix~\ref{appendix:score_velocity}). Setting $\gdiff_\tau=0$ recovers the deterministic ODE. The freedom to choose $\gdiff_\tau>0$ lets a deterministic generator be driven by an SDE, which is required by the posterior construction in Section~\ref{sec:methodology}.

\paragraph{Instance 1: stochastic interpolants.}
The stochastic interpolant (SI) framework \cite{albergo_stochastic_2023, chen_probabilistic_2024} takes the deterministic anchor to be the previous state, $\bm{a}_0=\bx_0=\state^{n-1}$, and embeds the noise along the path as a Brownian motion, $\sigpath_\tau\latentnoise = \gamma_\tau\bW_\tau$ with $\bW_\tau=\sqrt{\tau}\,\bz$, $\bz\sim\mathcal{N}(0,\bI)$. The interpolant becomes
\begin{align}\label{eq:stochastic_interpolant}
    \bx_\tau = \alpha_\tau\bx_0 + \beta_\tau\bx_1 + \gamma_\tau\bW_\tau,
\end{align}
with $\alpha_0=\beta_1=1$ and $\alpha_1=\beta_0=\gamma_1=0$, so that $\bx_{\tau=0}=\bx_0$ and $\bx_{\tau=1}\sim\conddist{\bx_1}{\bx_0}$. Its dynamics are
\begin{align}\label{eq:true_SI_SDE}
    \rd\bx_\tau = \underbrace{\left(\dot\alpha_\tau\bx_0 + \dot\beta_\tau\bx_1 + \dot\gamma_\tau\bW_\tau\right)}_{=:\,\bR_\tau(\bx_0,\bx_1)}\rd\tau + \gamma_\tau\,\rd\bW_\tau .
\end{align}
The model is an SDE
\begin{align}\label{eq:SI_SDE}
    \rd\bX_\tau = \drift_\theta(\bX_\tau,\bx_0,\tau)\,\rd\tau + \gamma_\tau\,\rd\bW_\tau, \qquad \bX_0=\bx_0,
\end{align}
whose drift is trained to match the conditional expectation $\drift(\bx,\bx_0,\tau) = \E[\bR_\tau(\bx_0,\bx_1)\mid\bX_\tau=\bx]$ by minimizing
\begin{align}\label{eq:SI_drift_loss}
    \mathcal{L}(\theta) = \int_0^1 \E\!\left[\norm{\drift_\theta(\bx_\tau,\bx_0,\tau) - \bR_\tau}_2^2\right]\rd\tau,
\end{align}
with the expectation over $\bx_0,\bx_1\sim\dist{\bx_0,\bx_1}$ and $\bz$, and $\tau\sim\mathcal{U}[0,1]$. No SDE solve is required during training. Here the SDE drift already includes the score contribution, i.e.\ $\drift_\theta = \fmvel_\tau + \tfrac12\gamma_\tau^2\genscore_\tau$ with $\gdiff_\tau=\gamma_\tau$ in Eq.~\eqref{eq:general_sde_family}, and the score is recovered as \cite{chen_probabilistic_2024}
\begin{align}\label{eq:SI_score}
    \genscore_\tau(\bx,\bx_0) = A_\tau\!\left[\beta_\tau\drift_\theta(\bx,\bx_0,\tau) - c_\tau(\bx,\bx_0)\right],
\end{align}
with $A_\tau = \left[\tau\gamma_\tau(\dot\beta_\tau\gamma_\tau - \beta_\tau\dot\gamma_\tau)\right]^{-1}$ and $c_\tau(\bx,\bx_0)=\dot\beta_\tau\bx + (\beta_\tau\dot\alpha_\tau - \dot\beta_\tau\alpha_\tau)\bx_0$.

\paragraph{Instance 2: flow matching.}
Flow matching (FM) \cite{lipman_flow_2023, liu_rectified_2022} takes the source to be Gaussian, $\bm{a}_0=\bm{0}$ and $\sigpath_\tau=\alpha_\tau$ acting on a latent $\latent\sim\mathcal{N}(0,\bI)$, with the previous state $\state^{n-1}$ entering only as a conditioning input to the network. The interpolant is the deterministic path
\begin{align}\label{eq:fm_interpolant}
    \bx_\tau = \alpha_\tau\latent + \beta_\tau\bx_1, \qquad \alpha_0=\beta_1=1,\quad \alpha_1=\beta_0=0,
\end{align}
e.g.\ $\alpha_\tau=1-\tau$, $\beta_\tau=\tau$ for rectified flow. The model is the ODE
\begin{align}\label{eq:fm_ode}
    \rd\bX_\tau = \fmvel_\theta(\bX_\tau,\state^{n-1},\tau)\,\rd\tau, \qquad \bX_0\sim\mathcal{N}(0,\bI),
\end{align}
whose velocity is trained to match $\fmvel_\tau(\bx) = \E[\dot\alpha_\tau\latent + \dot\beta_\tau\bx_1\mid\bX_\tau=\bx]$ through the conditional flow matching loss
\begin{align}\label{eq:fm_loss}
    \mathcal{L}(\theta) = \int_0^1 \E\!\left[\norm{\fmvel_\theta(\bx_\tau,\state^{n-1},\tau) - (\dot\alpha_\tau\latent + \dot\beta_\tau\bx_1)}_2^2\right]\rd\tau,
\end{align}
with the expectation over $\bx_1\sim\conddist{\bx_1}{\state^{n-1}}$ and $\latent$. As with SI, no ODE solve is needed during training, and the per-sample target equals the intractable marginal velocity in expectation \cite{lipman_flow_2023}. Specializing Eq.~\eqref{eq:general_score_velocity} to $\bm a_0=\bm 0$, $\sigpath_\tau=\alpha_\tau$ gives the score directly from the trained velocity,
\begin{align}\label{eq:fm_score}
    \genscore_\tau(\bx) = \frac{\beta_\tau\,\fmvel_\theta(\bx,\state^{n-1},\tau) - \dot\beta_\tau\,\bx}{\alpha_\tau\,(\dot\beta_\tau\alpha_\tau - \dot\alpha_\tau\beta_\tau)},
\end{align}
which for rectified flow reduces to $\genscore_\tau(\bx) = (\tau\,\fmvel_\theta - \bx)/(1-\tau)$.

In both instances the trained model provides a velocity (equivalently a score) for an affine Gaussian path, and Eq.~\eqref{eq:general_sde_family} turns it into an SDE with a diffusion coefficient of our choosing. This is the common interface on which the data assimilation method of Section~\ref{sec:methodology} operates.
